Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-09-04 writeups/proposal_presentation/proposal_presentation_sv.qmd
updated based on feedback

## Background
- Stochastic volatility models the variance of a financial instrument as a random variable

- Variety of Bayesian estimation strategies using classic Markov Chain Monte Carlo (MCMC) algorithms such as Gibbs and Metropolis Hastings

- Since then, more advanced algorithms such as Hamiltonian Monte Carlo has enabled the estimation of richer and more complicated models

- As the developments of these tools enable the estimation of more complicated models so do the risk of mistakes and the need to validate the output of our algorithms. How do we ensure the validity of our inferences across different estimation strategies?

<!-- What is the fundamental difference between convergence diagnostics, bayes factors with SBC? Check the fit of the model?-->

@@

- 3 key components
    + Kim, Sherphard and Chib (1998) SV model
    + Hamiltonian Monte Carlo
    + Simulation Based Calibration to validate MCMC computation

@@

@@

$$
\begin{aligned}
y_t =& \space \beta e^{h_t/2} \epsilon_t \\
h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
h_1 \sim& \space normal\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
\epsilon_t \sim &N(0,1) \space\space\space \eta_t \sim N(0,1)
\end{aligned}
$$

<!-- :::{.nonincremental} -->
- $y_t$ is mean corrected returns for equally spaced intervals $t$

- $\beta = exp(\mu / 2)$ constant scaling factor (set to 1)

- $h_t$ is log volatility where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ is a stationary process governed by the autoregressive parameter $\phi$ where $|\phi|<1$.

- $\mu$ is the "average" log volatility

<!-- ::: -->

## Kim Sherphard Chib (1998)
:::{.nonincremental}
- Estimation of states using a Kalman filter, however, this requires the model to be linear and Gaussian.

@@

## Kim Sherphard Chib (1998) Estimation strategy
Model is now linear but not Gaussian. KSC use a mixture of Gaussians to **approximate** the first 4 moments of the log chi squared distribution through moment matching. This is defined by:

@@

    + Runs physics simulation - imagine the parameter is a marble rolling in random directions inside a bowl

@@

Therefore, any deviation of the posterior samples from the prior distribution suggests an error - either in computation, code or overall analysis.

>>>>

# What is Stochastic Volatility?

## Log returns S&P 500
![]("returns.png")

## Stochastic Volatility (SV) model

$$
\begin{aligned}
y_t =& \space \beta e^{h_t/2} \epsilon_t \\
h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
h_1 \sim&  N\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
\epsilon_t \sim& N(0,1) \quad \eta_t \sim N(0,1) \\
\end{aligned}
$$

- $y_t$ is mean corrected returns for equally spaced intervals $t$

- $\beta = e^{\mu / 2}$ constant scaling factor (set to 1)

- $h_t$ is log volatility where $h_1$ is a draw from a stationary distribution

- State equation $h_{t+1}$ is a stationary process where $\mu$ is the "average" log volatility and determined by the autoregressive parameter $\phi$ where $|\phi|<1$.

@@

- Compare the Kim, Shephard and Chib (1998) estimation strategy with Hamiltonian Monte Carlo

- Simulation Based Calibration to validate MCMC computation

@@

$$
\begin{aligned}
y_t =& \space \beta e^{h_t/2} \epsilon_t \\
h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
h_1 \sim& N\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
\epsilon_t \sim& N(0,1) \quad \eta_t \sim N(0,1) \\
\end{aligned}
$$

@@

With priors:
$$
\begin{aligned}
\mu \sim& \space N(0, 10^2) \\
\sigma_{\eta}^2 \sim& \space IG(5/2, (0.01\times 5) / 2) \\
\phi^{\ast} \sim& \space beta(20, 1.5) \\
\phi =& 2\phi^{\ast} - 1
\end{aligned}
$$

## Kim Shephard Chib (1998)
:::{.nonincremental}
- Estimation of states using a standard Kalman filter, however, this requires the model to be linear and Gaussian.

@@

## Kim Shephard Chib (1998)
Model is now linear but not Gaussian. KSC use a mixture of Gaussians to **approximate** the first 4 moments of the log chi squared distribution through moment matching. This is defined by:

@@

    + Imagine the parameter is a marble rolling in random directions inside a bowl (negative log transformation of joint posterior)

@@

Therefore, if the distribution of rank statistics deviates from a discrete uniform distribution, there may be evidence that our analysis or computation is not returning the correct posteriors (not calibrated).

### 2023-09-04 writeups/research_proposal/propsal.qmd
updated based on feedback

---
title: "Comparison of MCMC algorithms in Stochastic Volatility Models"
format:
  pdf:
    fig-pos: H
    include-in-header:
        text: |
          \usepackage{lipsum}
          \usepackage{setspace}
          \onehalfspacing
          \linespread{1.5}
fontsize: 12pt
geometry: margin=1in
---

@@

## Research Goal
The objective of this research is to validate and compare the computational methods used to estimate Bayesian stochastic volatility models. Specifically, Simulation Based Calibration will be used to identify biases and errors arising from the results of Markov Chain Monte Carlo (MCMC) algorithms. The source of any miscalibration include, but are not limited to, estimation strategies, choice of sampler, model specifications and coding mistakes. This study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo for estimating these class of models. Results and diagnostics from these simulations will help identify strengths and limitations that arise from these approaches.

@@

# Background and Literature Review

@@

# 3) Methodology
The aim of this research is to use Simulation Based Calibration (SBC) to validate the inference of posterior samples generated by different MCMC algorithms in the context of stochastic volatility models. This will help identify any bias in computation and inconsistencies in model implementation and estimation strategies. Comparison of the SBC results between different computational approaches may help identify any strengths or shortcomings as well as any errors in model specification.

## 3.1) Stochastic Volatility Model
The univariate stochastic volatility model described by Kim, Shephard and Chib (1998) models the variance as a random variable following some latent stochastic process. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is its variance. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks.

@@

## 3.2) Estimation Strategy

@@

# Results
SBC for Hamiltonian Monte Carlo and the Gaussian approximation are performed with simulated dataset sizes of 1000 simulation iterations, 1000 observations, and post warmup/burnin samples of 999 (which gives 1000 possible rank statistics). The histogram is drawn with 20 bins and 50 observations in each bin which is the expected count for a discrete uniform distribution. This algorithm is summarised below:

@@

# Conclusion and next steps
There are two key extensions to the simulation results worth exploring.

@@

@@

# Appendix B: Conjugate posterior distributions

>>>>

---
title: "Comparison of MCMC algorithms in Stochastic Volatility Models"
author: 
    - "By Benjamin Wee" 
    - "Supervised by Catherine Forbes and Lauren Kennedy"
# abstract: "The objective of this research is to validate and compare the computational methods used to estimate Bayesian stochastic volatility models. Specifically, Simulation Based Calibration will be used to identify biases and errors arising from the results of Markov Chain Monte Carlo (MCMC) algorithms. The source of any miscalibration include, but are not limited to, estimation strategies, choice of sampler, model specifications and coding mistakes. This study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo for estimating these class of models. Results and diagnostics from these simulations will help identify strengths and limitations that arise from these approaches."
format:
  pdf:
    fig-pos: H
    include-in-header:
        text: |
          \usepackage{lipsum}
          \usepackage{setspace}
          \onehalfspacing
          \linespread{1.5}
fontsize: 12pt
geometry: margin=1in
mathspec: true
---

@@

## 1. Research Goal
The objective of this research is to validate and compare the computational methods used to estimate Bayesian stochastic volatility models. Specifically, Simulation Based Calibration will be used to identify biases and errors arising from the results of Markov Chain Monte Carlo (MCMC) algorithms. The source of any miscalibration include, but are not limited to, estimation strategies, choice of sampler, model specifications and coding mistakes. This study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo for estimating these class of models. Results and diagnostics from these simulations will help identify strengths and limitations that arise from these approaches.

@@

# 2. Background and Literature Review

@@

# 3. Methodology
The aim of this research is to use Simulation Based Calibration (SBC) to validate the inference of posterior samples generated by different MCMC algorithms in the context of stochastic volatility models. This will help identify any bias in computation and inconsistencies in model implementation and estimation strategies. Comparison of the SBC results between different computational approaches may help identify any strengths or shortcomings as well as any errors in model specification.

## 3.1 Stochastic Volatility Model
The univariate stochastic volatility model described by Kim, Shephard and Chib (1998) models the variance as a random variable following some latent stochastic process. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is its variance. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks.

@@

## 3.2 Estimation Strategy

@@

# 4. Results
SBC for Hamiltonian Monte Carlo and the Gaussian approximation are performed with simulated dataset sizes of 1000 simulation iterations, 1000 observations, and post warmup/burnin samples of 999 (which gives 1000 possible rank statistics). The histogram is drawn with 20 bins and 50 observations in each bin which is the expected count for a discrete uniform distribution. This algorithm is summarised below:

@@

# 5. Conclusion and next steps
There are two key extensions to the simulation results worth exploring.

@@

{{< pagebreak >}}

@@

| $\omega$ | $Pr(\omega = i)$ | $m_i$ | $\nu^2_i$ |
|------|------|------|------|
| 1    | 0.00730    | -10.12999   | 5.79596 | 
| 2    | 0.10556    | -3.97281    | 2.61369 | 
| 3    | 0.00002    | -8.56686    | 5.17950 | 
| 4    | 0.04395    | 2.77786    | 0.16735 | 
| 5    | 0.34001    | 0.61942    | 0.64009 | 
| 6    | 0.24566    | 1.79518    | 0.34023 | 
| 7    | 0.25750    | -1.08819    | 1.26261 |

{{< pagebreak >}}

# Appendix B: Sampling from mixture of Gaussians

## Conjugate posterior distributions

**Sampling** $\boldsymbol{\sigma_{\eta}^2}$

Inverse gamma conjugate posterior distribution:

$$
\sigma^2_{\eta} | y,h,\phi,\mu \sim IG \Bigl\{\frac{n+\sigma_r}{2}, \frac{0.05 + (h_1 - \mu)^2 (1 - \phi^2) + \sum_{t=1}^{n-1}((h_{t+1} - \mu) - \phi(h_t - \mu))^2}{2}\Bigr\}
$$

**Sampling** $\boldsymbol{\mu}$

Gaussian conjugate posterior distribution:

$$
\mu | h,\phi,\sigma^2_{\eta}  \sim N(\hat{\mu}, \sigma^2_{\mu})
$$

Where

$$
\begin{aligned}
\hat{\mu} &= \sigma^2_{\mu} \Bigl\{\frac{(1-\phi^2)}{\sigma_{\eta}^2}h_1 +\frac{(1-\phi^2)}{\sigma_{\eta}^2} \sum_{t=1}^{n-1} (h_{t+1} - \phi h_t)\Bigr\} \\
\sigma^2_{\mu} &= \sigma^2_{\eta} \{(n-1)(1-\phi)^2 + (1-\phi^2)\}^{-1}
\end{aligned}
$$

**Sampling** $\boldsymbol{\phi}$

Metropolis Hastings accept/reject procedure:

1) Generate proposal $\phi^\ast$ from $N(\hat{\phi}, V_{\phi})$ where $\hat{\phi} = \frac{\sum_{t=1}^{n-1} (h_{t+1} - \mu)(h_t - \mu)}{\sum_{t=1}^{n-1} (h_t - \mu)^2}$ and $V_{\phi} = \sigma^2_{\eta} \{\sum_{t=1}^{n-1} (h_t - \mu)^2\}^{-1}$

2) Accept proposal as $\phi^{(i)}$ with probability $e^{\{g(\phi^\ast) - g(\phi^{(i-1)}\}}$ such that $g(\phi) = log (\pi (\phi)) - \frac {(h_t - \mu)^2 (1-\phi^2)}{2 \sigma_{\eta}^2} + \frac{1}{2} log (1-\phi^2)$
